Flare Rates and Rotation Periods of 5700 Stars Younger than 300 Myr

### 2023-10-31 src/tex/ms.tex
rossby # histograms

% Define document class
\documentclass[twocolumn]{aastex631}
\usepackage{showyourwork}

@@

% Title
\title{Evidence of Magnetic Saturation from Stellar Flare Rates}

@@

% Abstract with filler text
\begin{abstract}
    Photometric exoplanet detection missions have revolutionized our understanding
    of stellar activity. Stellar flares are short-duration high-energy radiation
    outbursts which can be readily seen across young (< 800\,Myr) stars and even
    in high counts on mature M dwarfs. Here, we present the results of measuring
    stellar flare rates for very young stars, with ages $4.5 - 200$\,Myr right
    after they formed to understand the early evolution of magnetic activity. We
    have selected a sample of 5,700 stars which belong to 26 associations,
    nearby young moving groups, or clusters. We identified $\sim 29,000$ flares
    using the \texttt{stella} machine-learning algorithm. We measured the flare-
    frequency distribution slopes as a function of age and stellar effective
    temperature.
\end{abstract}

@@

\subsection{Measuring Rotation Periods}

\cite{seligman22} demonstrated that stars with low Rossby numbers $R_0 < 0.13$ have
shallower flare frequency distribution slopes, indicative of more high energy flares
originating from these sources. In addition to understanding flare statistics across
young ages, we aim to expand this sample by measuring the rotation periods, $P_\textrm{rot}$,
for the stars in our sample. To do this, we used \texttt{michael}\footnote{\url{https://github.com/ojhall94/michael}},
an open-source Python package that robustly measures $P_\textrm{rot}$ using a combination
traditional Lomb-Scargle periodograms and wavelet transformations (Hall et al. submitted).
\texttt{michael} measures $P_\textrm{rot}$ using the \texttt{eleanor} package, which
extracts light curves from the TESS Full-Frame Images \citep[FFIs;][]{feinstein19}.
We ran \texttt{michael} on all stars from which flares were identified. The estimated
$P_\textrm{rot}$ were vetted by-eye, from the \texttt{michael} diagnostic plots. In
total, we robustly measured \textcolor{red}{n} $P_\textrm{rot}$ across our sample of
3,983 stars.

@@

\begin{figure*}[ht!]
    \script{mcmc_results.py}
    \begin{centering}
        \includegraphics[width=3\textwidth]{figures/mcmc_results.pdf}
        \caption{
            Measured flare-frequency distribution slopes, $\alpha$, as a function
            of stellar effective temperature, $T_\textrm{eff}$ and age.
        }
        \label{fig:mcmc_results}
    \end{centering}
\end{figure*}

@@

@@

We measured the flare rates of planet hosting stars $< 300$\,Myr, comparable to the ages of our primary sample.
We followed the methods outlined in Section~\ref{sec:methods} to detect and vet flares. For each star, we created
an equivalent sample with respect to both age and $T_\textrm{eff}$ to compare the flare rates too. We considered
stars with ages $\pm 30$\,Myr of the planet hosting star and $T_\textrm{eff} \pm 1000$\,K. We calculated the overall
flare rate as Equation~\ref{eq:fr}. We report the flare rates of host star and the comparable sample in Table~\ref{tab:yprates}.
For the comparable sample, we report the median flare rate, and the lower 16\textsuperscript{th} and upper 84\textsuperscript{th}
percentiles.

@@

>>>>

% Define document class
\documentclass[twocolumn]{aastex631}
\usepackage{showyourwork}
\usepackage{rsfso}

@@

% Title
\title{Flare Rates and Rotation Periods of 5700 Stars Younger than 300 Myr}

@@

% Abstract with filler text
\begin{abstract}
    Photometric exoplanet detection missions have revolutionized our understanding
    of stellar activity. Stellar flares are short-duration high-energy radiation
    outbursts which can be readily seen across young (< 800\,Myr) stars and even
    in high counts on mature M dwarfs. Here, we present the results of measuring
    stellar flare rates for very young stars, with ages $4.5 - 200$\,Myr right
    after they formed to understand the early evolution of magnetic activity. We
    have selected a sample of 5,700 stars which belong to 26 associations,
    nearby young moving groups, or clusters. We identified $\sim 29,000$ flares
    using the \texttt{stella} machine-learning algorithm. We measured the flare-
    frequency distribution slopes as a function of age and stellar effective
    temperature. Finally, we measure the flare rates of young planet-hosting stars and
    compare flare rates to a larger sample of stars with similar ages and $T_\textrm{eff}$.
    We find that the majority of planet-hosting stars have lower flare rates, with the exception of
    HIP 67522, AU Mic, DS Tuc, and TOI 451.
\end{abstract}

@@

\subsection{Measuring Rotation Periods}\label{subsec:prot}

\cite{seligman22} demonstrated that stars with low Rossby numbers $R_0 < 0.13$ have
shallower flare frequency distribution slopes, indicative of more high energy flares
originating from these sources. In addition to understanding flare statistics across
young ages, we aim to expand this sample by measuring the rotation periods, $P_\textrm{rot}$,
for the stars in our sample. To do this, we used \texttt{michael}\footnote{\url{https://github.com/ojhall94/michael}},
an open-source Python package that robustly measures $P_\textrm{rot}$ using a combination
traditional Lomb-Scargle periodograms and wavelet transformations (Hall et al. submitted).
\texttt{michael} measures $P_\textrm{rot}$ using the \texttt{eleanor} package, which
extracts light curves from the TESS Full-Frame Images \citep[FFIs;][]{feinstein19}.
We ran \texttt{michael} on all stars from which flares were identified. The estimated
$P_\textrm{rot}$ were vetted by-eye, from the \texttt{michael} diagnostic plots. In
total, we robustly measured 1,269 $P_\textrm{rot}$ across our sample of
3,983 stars, which is $\sim 50\%$.

@@

\begin{figure*}[ht!]
    \script{mcmc_results.py}
    \begin{centering}
        \includegraphics[width=3\textwidth, trim={0 0 30mm 0}, clip]{figures/mcmc_results.pdf}
        \caption{
            Measured flare-frequency distribution slopes, $\alpha$, as a function
            of stellar effective temperature, $T_\textrm{eff}$ and age.
        }
        \label{fig:mcmc_results}
    \end{centering}
\end{figure*}

@@

\section{Flare Rates Compared to P$_{rot}$}

\begin{figure}[ht!]
    \script{prot_histograms.py}
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/prot_histograms.pdf}
        \caption{
            Rotation periods and flare rates for M, K, and G stars (top to bottom).
        }
        \label{fig:prot_histograms}
    \end{centering}
\end{figure}

@@

We measured the flare rates of planet hosting stars $< 300$\,Myr, comparable to the ages of our primary sample.
We followed the methods outlined in Section~\ref{sec:methods} to detect and vet flares. For each star, we created
an equivalent sample with respect to both age and $T_\textrm{eff}$ to compare the flare rates too. We considered
stars with ages $\pm 30$\,Myr of the planet hosting star and $T_\textrm{eff} \pm 1000$\,K. We calculate the overall
flare rate as described in \cite{Feinstein22}:

\begin{equation}
  \mathcal{R} = \frac{1}{\tau_\textrm{obs}} \left( \sum_{i=1}^{N} p_i \right)
\end{equation}

where $\mathcal{R}$ is the flare rate in day$^{-1}$, $\tau_\textrm{obs}$ is the total amount of time a target was
observed with TESS, and $p_i$ is the probability that flare $i$ is a true flare as assigned by \texttt{stella}. We present
the flare rates of planet-hosting stars and a comparable sample of stars in Figure~\ref{fig:yp_rates} and report the measured
rates in Table~\ref{tab:yp_rates}. For the comparable sample, we report the median flare rate, and the lower 16\textsuperscript{th}
and upper 84\textsuperscript{th} percentiles.

@@

X-ray luminosity surveys of stars have revealed a saturation limit with respect to
the star's rotation period. Namely, there is no evolution in $L_X/L_\textrm{bol}$ for
stars with $P_\textrm{rot} < 10$\,days \citep{Pizzolato03}.

Here, we explore if flare rates evolve as a function of rotation period. We used our
measured $P_\textrm{rot}$ as described in Section~\ref{subsec:prot}. Due to TESS's orbit,
we limit our search of $P_\textrm{rot}$ to $< 20$\,days. This should not bias our analysis as
young stars are known to rotate on short timescales \citep[e.g.][]{west15, popinchalk23}. We
present our results in Figure~\ref{fig:prot}.
